Genome-wide changes in expression during inflorescence architecture development

### 2015-10-16 ms.Rmd
final commit before changing gene expression cutoff

Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that are increasingly expressed between RM and SM. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go]. Reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg]. These results suggest that an increase in active cytokinin is associated with grain yield via an effect on panicle branching, but the function of *LABA1* suggests the opposite. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. No reports were found of *LOGL1* function, or the function of the gene annotated as a cytokinin dehydrogenase (*LOC_Os05g31040*) that is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), but the expression pattern of *LABA1* and these two genes suggests increasing cytokinin activation and decreasing inactivation from RM to SM. It is not currently clear how such a mechanism might influence meristem identity or panicle branching at the molecular level. [**Check for at/tomato papers**]. There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex.

@@

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. The G1-Like (G1L/ALOG) genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj]. Only three were detected in the dataset. Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*. Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55* [@Yoshida:2013ff]. *MADS22* and *MADS55* are also highly expresed in the RM, and the third rice *STMADS11*-like gene, *MADS47*, is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).

@@

The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. **Downstream targets of WFP**. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].

@@

The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by @Prunet:2013ki). In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**]. [**Add info about tomato MADS**].

@@

**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**. Neighbour-joining similarity tree of putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*. Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers. The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do]. For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes. Clades are labelled following the nomenclature presented by @Arora:2007fa.

>>>>

Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM. In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**]. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go]. Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg]. These results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LABA1* gene provides evidence for an opposite function. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. The function of *LOGL1* has not been reported. There is also an uncharacterized gene, *LOC_Os05g31040*, which is annotated as a cytokinin dehydrogenase, present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`). The expression patterns of *LABA1*, *LOGL1* and *LOC_Os05g31040* suggest increasing cytokinin activation and decreasing inactivation from RM to SM. Such a mechanism would indicate a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level. Genes that may be involved in cytokinin signalling are also present other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.

@@

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. The G1-Like (G1L/ALOG) genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj]. Only three were detected in the dataset. Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*. Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, is not reported [@Yoshida:2013ff]. *MADS22* and *MADS55* are also enriched in the RM, and *MADS47* is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).

@@

| **Shorten MADS section and insert here. MADS DEG figure goes to SI?**
| **GRAS TFs (SHR1/SCR)?**

The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc]. Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf]. WFP also binds to the promoters of *LOG1*, *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].

| **[Re]move following MADS section and insert section on Homeobox. Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007)**

@@

The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki]. In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**]. [**Add info about tomato MADS**].

@@

**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**. Differential expression of putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*. The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences. Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers. The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the adjusted *p*-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do]. For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes. Clades are labelled following the nomenclature presented by @Arora:2007fa.
